Between Subgraph Isomorphism and Maximum Common Subgraph

### 2016-09-11 paper/paper.tex
We don't need F without CBJ
Complements to the algorithm
Committing from 36,000ft

@@

In this work we discuss an intermediate problem, where we must map all but $k$ vertices of the
pattern graph into the target. We show that if $k$ is reasonably small (say, between 1 and 5), then
weakened forms of degree and path based filterings are still effective in pruning the initial search
space and in providing additional constraints respectively. We then show that combining these
techniques leads to a practical algorithm which can scale to work with the families of graphs
commonly used to benchmark subgraph isomorphism algorithms: depending upon the benchmark family, in
a substantial portion of ??; in ?? more cases, we can at least obtain an upper bound.

@@

The \emph{complement} of a (simple) graph $P$ is the graph $\bar{P}~$.  Adjacent distinct vertices in $P$ are non-adjacent in $\bar{P}$ while non-adjacent vertices in $P$ are adjacent in $\bar{P}$.  The \emph{loop complement} of a graph $P$ is the graph $\loopcomp{P}$.  The constructcion of $\loopcomp{P}$ is similar to that of $\bar{P}~$, however whenever a vertex has a loop in $P$ it does not in $\loopcomp{P}$ and vice versa.  When solving an induced subgraph isomorphism problem we make use of the loop complement; we care about mapping vertices with loops to vertices with loops, and edges without loops to vertices without loops.  (Although when we filter by paths we make use of the complement rather than the loop complement.)  The following proposition follows directly from the definitions.

@@

For both subgraph isomorphism and maximum common subgraph, constraint-based models are the best
approach (although constraint programming toolkits are not used, to facilitate better memory layouts
and propagation queues). For both problems, we create a variable for each vertex in the pattern
graph (the smaller graph, in the case of maximum common subgraph), with domains ranging over the
vertices in the target graph (the larger graph). For maximum common subgraph, each domain is given
an additional $\bot$ value, meaning ``unmapped''.

@@

In our algorithm we make use of \cref{prop:comp} by seeking an $i$ which simultaneously satisfies $P
\rightarrowtail T$ and $\bar{P} \rightarrowtail \bar{T}$ when searching for an induced subgraph
isomorphism.  Further, \cref{prop:comp} tells us that every induced subgraph isomorphism is
equivalent to simultaneous non-induced subgraph isomorphisms both $P$ and $T$ as well as their
complements.

All different.

For subgraph isomorphism, we can also do domain filtering at the top of search. We discuss this in
\cref{section:degreefiltering}. Additional constraints, discuss this in \cref{section:pathfiltering}.

@@

\begin{figure}
    \includegraphics*{gen-graph-ids.pdf}
    \caption{The amount of domain reduction achieved for the induced problem, with increasing values
    of $k$. The $y$ value shows for how many instances we may reduce the domain size to be at most
    the $x$ proportion of the size the domains would be for the maximum common subgraph
    problem.}\label{figure:ids}
\end{figure}

@@

To allow for fast propagation, we create supplemental graphs for paths of length 2, looking at
counts of 1, 2, and 3 in the target graph (and so counts of $1 + k$ up to $3 + k$ in the pattern).
We then investigate whether this leads to new constraints being generated. By an assignment, we mean
considering mapping a pattern vertex $p$ to a target vertex $t$ (and not $\bot$) which does not
violate any loop constraints. An assignment pair is two assignments with distinct $p$ and distinct
$t$, which we say is permitted if it does not violate any adjacency constraint. We define the
permitted assignment pair ratio to be the proportion of assignment pairs which are permitted.
Finally, we scatter plot the permitted assignment pair ratio with and without supplemental
graphs\footnote{Because of the large sizes of the domains, we randomly sample one million pairs
rather than considering every pair. In some cases, we have nearly a thousand domains, each with
nearly ten thousand values---a complete quadratic calculation involving even a trivial arithmetic
operation on this would take many hours.}. For $k = 0$, we see many points above the $x-y$ diagonal,
which shows that for many instances, we are able to create a substantial number of new constraints
at the top of search; on the other hand, there are also points on the diagonal, which shows that
sometimes this technique provides no benefit. (Occasionally, points fall below the $x-y$ diagonal.
This is because the use of neighbourhood degree sequence reasoning on supplemental graphs can also
lead to increased domain filtering, which could in turn eliminate a higher proportion of forbidden
than permitted assignment pairs.) For $k = 1$ and $k = 2$, the proportion of points above the
diagonal diminishes, but we are still able to create new constraints for many instances. By the time
$k = 3$, most of the benefit is disappearing---although sometimes we are still able to make a
difference, and bear in mind that sometimes adding just one new constrained pair can vastly reduce
the search space.  A closer inspection of the raw data shows that for $k \ge 3$, nearly all of the
instances showing a strong improvement are from the ``LV'' family of graphs.

@@

\begin{algorithm*}[tb]
\DontPrintSemicolon
\begin{multicols}{2}
\nl $\FuncSty{klessSubgraphIsomorphism}$ (\\\hspace*{2em}Graph $\mathcal{P}$, Graph $\mathcal{T}$, Int $k$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \lIf{$\left|\V(\mathcal{P})\right| + k > \left|\V(\mathcal{T})\right|$}{$\KwSty{return}~\KwSty{false}$\label{line:enough}}
    \nl $L \gets \big[ (\mathcal{P},\ \mathcal{T}), (\shortoverline{\mathcal{P}},
        \shortoverline{\mathcal{T}}),$
        \\\hspace*{1.0em}$(\mathcal{P}^{1 + k,2},\ \mathcal{T}^{1,2}), (\mathcal{P}^{2 + k,2},\
    \mathcal{T}^{2,2}), (\mathcal{P}^{3 + k,2},\ \mathcal{T}^{3,2}) \big]$ \;
    \nl \ForEach{$v \in \V(\mathcal{P})$}{
        \nl $D_v \gets \V(\mathcal{T})$\;
        \nl \ForEach{$(P,\,T) \in L$}{
            \nl $D_v \gets \{ w \in D_v : $ \\
            \hspace*{4em}$v \sim_P v \Rightarrow w \sim_T w~\wedge$ \\
            \hspace*{4em}$\nds_P(v) \lesspreceq{k} \nds_T(w)\}$ \;
        }
        \nl $D_v \gets D_v \cup k~\textnormal{distinct wildcard values}$
    }
    \nl \lIf{\FuncSty{propagate}(D)}{$\KwSty{return}~\FuncSty{search}(L, D, k)$}
    \nl \lElse{$\KwSty{return}~\KwSty{false}$}
}
\bigskip
\nl $\FuncSty{search}$ (GraphPairs $L$, Domains $D$, Int $k$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \lIf{$D = \emptyset$}{$\KwSty{return}~\KwSty{true}$}
    \nl $D_v \gets \textnormal{the smallest domain in}~D$ \;
    \nl \ForEach {$v' \in D_v~\textnormal{ordered by static degree in}~\mathcal{T}$\label{line:parforeach}}{
        \nl \If{$v'$ \textnormal{is not the first wildcard we have tried}}{
            \nl \KwSty{continue} \;
        }
        \nl $D' \gets \FuncSty{clone}(D)$ \;
        \nl $D'_v \gets \{ v' \} $ \;
        \nl \If{\FuncSty{propagate}(D')}{
            \nl \If{$\FuncSty{search}(L, \{ D \in D' : \left|D\right| > 1 \}, k)$}{
                \nl $\KwSty{return}~\KwSty{true}$ \;
            }
        }
    }
}
\bigskip
\nl $\FuncSty{propagate}$ (GraphPairs $L$, Domains $D$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \While{\KwSty{true}}{
        \nl \If{\textnormal{no effectively-unit domains (treating all wildcard values as a single
        value) remain}}{
            \nl \If{$\KwSty{not}~\FuncSty{allDifferent}(D)$}{
                \nl $\KwSty{return}~\KwSty{false}$ \;
            }
        }
        \nl \If{\textnormal{no effectively-unit domains remain}}{
            \nl $\KwSty{return}~\KwSty{true}$ \;
        }
        \nl $D_v \gets \textnormal{an effectively-unit domain from}~D$ \;
        \nl $v' \gets $ the single value in $D_v$, or an arbitrary wildcard value if there is more
        than one \;
        \nl \ForEach{$D_w \in D \setminus \{ D_v \}$}{
            \nl $D_w \gets D_w \setminus \{ v' \}$ \label{line:removev} \;
            \nl \ForEach{$(P,\,T) \in L$}{
                \nl \If{$v \sim_P w$}{
                    \nl $D_w \gets D_w \cap \left(\N_T(v') \cup \textnormal{wildcards}\right)$
                }
            }
            \nl \lIf{$D_w = \emptyset$}{$\KwSty{return}~\KwSty{false}$}
        }
    }
}
\bigskip
\nl $\FuncSty{allDifferent}$ (Domains $D$) $\rightarrow$ Bool \;
\nl \Begin{
\nl $(F,\,H,\,A,\,n) \gets (\emptyset,\,\emptyset,\,\emptyset,\,0)$ \;
\nl \ForEach{$D_v \in D$ \textnormal{from smallest to largest\label{line:eachdomain}}}{
    \nl $F \gets F \cup \{ v \}$ \;
    \nl $D_v \gets D_v \setminus H$ \label{line:elimhall} \;
    \nl $(A,\,n) \gets (A \cup D_v,\,n + 1)$ \label{line:acc} \;
    \nl \lIf{$D_v = \emptyset~\KwSty{or}~\left|A\right| < n$}{$\KwSty{return}~\KwSty{false}$\label{line:failhall}}
    \nl \lIf{$\left|A\right| = n$}{$(H,\,A,\,n) \gets (H \cup A,\,\emptyset,\,0)$\label{line:hall}}
}
\nl $\KwSty{return}~\KwSty{true}$ \;
}
\end{multicols}
\vspace*{0.4cm}% bug in algorithm2e + multicol: incorrect space left above the vertical line
\caption{An algorithm for the $k$-less subgraph isomorphism problem.}
\label{algorithm:sip}
\end{algorithm*}

@@

\begin{figure}
    \includegraphics*{gen-graph-runtimes.pdf}
    \caption{The cumulative number of instances solved over time, for the induced and non-induced
    problems, with different values of $k$. We also show the results of iteratively decreasing $k$
    until a solution is found, and in the induced case, the performance of a leading maximum common
    subgraph algorithm.}\label{figure:runtimes}
\end{figure}

@@

\begin{figure}
    \includegraphics*{gen-graph-which-k-by-family.pdf}
    \caption{The proportion of instances, in different families, which become satisfiable for
    increasing values of $k$. The larger instances are those where we can prove unsatisfiability for
    $k = 5$, whilst the gap between the top of the bar and the top of the graph is the fraction of
    instances where a timeout was reached for at least one value of $k$.}\label{figure:which-k}
\end{figure}

@@

% \section*{Acknowledgements}
% 
% Thanks to Christine, Samba, Patrick.

>>>>

\crefname{algocf}{Algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}

@@

In this work we discuss an intermediate problem, where we must map all but $k$ vertices of the
pattern graph into the target. We show that if $k$ is reasonably small (say, between 1 and 5), then
weakened forms of degree and path based filterings are still effective in pruning the initial search
space and in providing additional constraints respectively. We then show that combining these
techniques leads to a practical algorithm which can scale to work with the families of graphs
commonly used to benchmark subgraph isomorphism algorithms: depending upon the benchmark family, in
a substantial portion of ??; in ?? more cases, we can at least obtain an upper bound. This is a
significant improvement over state of the art maximum common subgraph solvers, which cannot even fit
many of these instances in RAM.

@@

The \emph{complement} of a (simple) graph $P$ is the graph $\shortoverline{P}~$.  Adjacent distinct
vertices in $P$ are non-adjacent in $\shortoverline{P}$ while non-adjacent vertices in $P$ are
adjacent in $\shortoverline{P}$.  The \emph{loop complement} of a graph $P$ is the graph
$\loopcomp{P}$.  The construction of $\loopcomp{P}$ is similar to that of $\bar{P}~$, however
whenever a vertex has a loop in $P$ it does not in $\loopcomp{P}$ and vice versa.  When solving an
induced subgraph isomorphism problem we make use of the loop complement; we care about mapping
vertices with loops to vertices with loops, and vertices without loops to vertices without loops.
(Although when we filter by paths we make use of the complement rather than the loop complement.)
The following proposition follows directly from the definitions.

@@

For both subgraph isomorphism and maximum common subgraph, constraint programming is the best known
approach\footnote{Note that toolkits are not used, to facilitate better memory layouts and
propagation queues, although the algorithms are heavily inspired by the constraint programming
literature}. For both problems, we create a variable for each vertex in the pattern graph (the
smaller graph, in the case of maximum common subgraph), with domains ranging over the vertices in
the target graph (the larger graph). For maximum common subgraph, each domain is given an additional
$\bot$ value, meaning ``unmapped''.

@@

In our algorithm we make use of \cref{prop:comp} by seeking an $i$ which simultaneously satisfies $P
\rightarrowtail T$ and $\loopcomp{P} \rightarrowtail \loopcomp{T}$ when searching for an induced subgraph
isomorphism.  Further, \cref{prop:comp} tells us that every induced subgraph isomorphism is
equivalent to simultaneous non-induced subgraph isomorphisms both $P$ and $T$ as well as their
complements.

?? All different.

For subgraph isomorphism, we can do domain filtering at the top of search, to reduce the initial
sizes of domains---we discuss this in \cref{section:degreefiltering}. Subgraph isomorphism also
allows us to generate additional implied constraints, which we discuss in
\cref{section:pathfiltering}. Neither of these techniques are valid for the maximum common subgraph
problem.

@@

Some other forms do not translate. For example, another invariant which can be effective on regular
graphs is to count the number of neighbours of a vertex which are present in a triangle. Removing a
single vertex can alter this count arbitrarily, so we cannot make use of this fact.

\begin{figure}
    \includegraphics*{gen-graph-ids.pdf}
    \caption{The amount of domain reduction achieved for the induced problem, with increasing values
    of $k$. The $y$ value shows for how many instances we may reduce the initial search space to be
    at most the $x$ proportion of the size the search space would be for the maximum common subgraph
    problem.}\label{figure:ids}
\end{figure}

@@

Unlike in conventional subgraph isomorphism, we cannot extend this filtering to look at paths of
length three: removing a single vertex can delete arbitrarily many such paths. We could count paths
of any length which are vertex disjoint, although calculating this is too expensive to be useful in
practice.

To allow for fast propagation, rather than calculating paths dynamically, we follow the approach of
\citet{DBLP:conf/cp/McCreeshP15} and construct supplemental graphs. We use paths of length 2,
looking at counts of 1, 2, and 3 in the target graph (and so counts of $1 + k$ up to $3 + k$ in the
pattern).  We then investigate whether this leads to new constraints being generated.

By an assignment, we mean considering mapping a pattern vertex $p$ to a target vertex $t$ (and not
$\bot$) which does not violate any loop constraints. An assignment pair is two assignments with
distinct $p$ and distinct $t$, which we say is permitted if it does not violate any adjacency
constraint. We define the permitted assignment pair ratio to be the proportion of assignment pairs
which are permitted.  Finally, we scatter plot the permitted assignment pair ratio with and without
supplemental graphs\footnote{Because of the large sizes of the domains, we randomly sample one
million pairs rather than considering every pair. In some cases, we have nearly a thousand domains,
each with nearly ten thousand values---a complete quadratic calculation involving even a trivial
arithmetic operation on this would take many hours.}.

For $k = 0$, we see many points above the $x-y$ diagonal, which shows that for many instances, we
are able to create a substantial number of new constraints at the top of search; on the other hand,
there are also points on the diagonal, which shows that sometimes this technique provides no
benefit. (Occasionally, points fall below the $x-y$ diagonal.  This is because the use of
neighbourhood degree sequence reasoning on supplemental graphs can also lead to increased domain
filtering, which could in turn eliminate a higher proportion of forbidden than permitted assignment
pairs.) For $k = 1$ and $k = 2$, the proportion of points above the diagonal diminishes, but we are
still able to create new constraints for many instances. By the time $k = 3$, most of the benefit is
disappearing---although sometimes we are still able to make a difference, and bear in mind that
sometimes adding just one new constrained pair can vastly reduce the search space.  A closer
inspection of the raw data shows that for $k \ge 3$, nearly all of the instances showing a strong
improvement are from the ``LV'' family of graphs.

@@

\begin{algorithm*}[tb]
\DontPrintSemicolon
\begin{multicols}{2}
\nl $\FuncSty{klessSubgraphIsomorphism}$ (\\\hspace*{2em}Graph $\mathcal{P}$, Graph $\mathcal{T}$, Int $k$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \lIf{$\left|\V(\mathcal{P})\right| + k > \left|\V(\mathcal{T})\right|$}{$\KwSty{return}~\KwSty{false}$\label{line:enough}}
    \nl $L \gets \big[ (\mathcal{P},\ \mathcal{T}), (\loopcomp{\mathcal{P}},
        \loopcomp{\mathcal{T}}) \textnormal{~only if we want induced},$
        \\\hspace*{1.0em}$(\mathcal{P}^{1 + k,2},\ \mathcal{T}^{1,2}), (\mathcal{P}^{2 + k,2},\
    \mathcal{T}^{2,2}), (\mathcal{P}^{3 + k,2},\ \mathcal{T}^{3,2}) \big]$ \;
    \nl \ForEach{$v \in \V(\mathcal{P})$}{
        \nl $D_v \gets \V(\mathcal{T})$\;
        \nl \ForEach{$(P,\,T) \in L$}{
            \nl $D_v \gets \{ w \in D_v : $ \\
            \hspace*{4em}$v \sim_P v \Rightarrow w \sim_T w~\wedge$ \\
            \hspace*{4em}$\nds_P(v) \lesspreceq{k} \nds_T(w)\}$ \;
        }
        \nl $D_v \gets D_v \cup k~\textnormal{distinct wildcard values}$
    }
    \nl \lIf{\FuncSty{propagate}(D)}{$\KwSty{return}~\FuncSty{search}(L, D, k)$}
    \nl \lElse{$\KwSty{return}~\KwSty{false}$}
}
\bigskip
\nl $\FuncSty{search}$ (GraphPairs $L$, Domains $D$, Int $k$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \lIf{$D = \emptyset$}{$\KwSty{return}~\KwSty{true}$}
    \nl $D_v \gets \textnormal{the smallest domain in}~D$ \;
    \nl \ForEach {$v' \in D_v~\textnormal{ordered by static degree in}~\mathcal{T}$\label{line:parforeach}}{
        \nl \If{$v'$ \textnormal{is not the first wildcard we have tried}}{
            \nl \KwSty{continue} \;
        }
        \nl $D' \gets \FuncSty{clone}(D)$ \;
        \nl $D'_v \gets \{ v' \} $ \;
        \nl \If{\FuncSty{propagate}(D')}{
            \nl \If{$\FuncSty{search}(L, \{ D \in D' : \left|D\right| > 1 \}, k)$}{
                \nl $\KwSty{return}~\KwSty{true}$ \;
            }
        }
    }
}
\bigskip
\nl $\FuncSty{propagate}$ (GraphPairs $L$, Domains $D$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \While{\KwSty{true}}{
        \nl \If{\textnormal{no effectively-unit domains (treating all wildcard~values as a single
        value) remain}}{
            \nl \If{$\KwSty{not}~\FuncSty{allDifferent}(D)$}{
                \nl $\KwSty{return}~\KwSty{false}$ \;
            }
        }
        \nl \If{\textnormal{no effectively-unit domains remain}}{
            \nl $\KwSty{return}~\KwSty{true}$ \;
        }
        \nl $D_v \gets \textnormal{an effectively-unit domain from}~D$ \;
        \nl $v' \gets $ the single value in $D_v$, or an arbitrary wildcard~value if there is more
        than one \;
        \nl \ForEach{$D_w \in D \setminus \{ D_v \}$}{
            \nl $D_w \gets D_w \setminus \{ v' \}$ \label{line:removev} \;
            \nl \ForEach{$(P,\,T) \in L$}{
                \nl \If{$v \sim_P w$}{
                    \nl $D_w \gets D_w \cap \left(\N_T(v') \cup \textnormal{wildcards}\right)$
                }
            }
            \nl \lIf{$D_w = \emptyset$}{$\KwSty{return}~\KwSty{false}$}
        }
    }
}
\bigskip
\nl $\FuncSty{allDifferent}$ (Domains $D$) $\rightarrow$ Bool \;
\nl \Begin{
\nl $(H,\,A,\,n) \gets (\emptyset,\,\emptyset,\,\emptyset,\,0)$ \;
\nl \ForEach{$D_v \in D$ \textnormal{from smallest to largest\label{line:eachdomain}}}{
    \nl $D_v \gets D_v \setminus H$ \label{line:elimhall} \;
    \nl $(A,\,n) \gets (A \cup D_v,\,n + 1)$ \label{line:acc} \;
    \nl \lIf{$D_v = \emptyset~\KwSty{or}~\left|A\right| < n$}{$\KwSty{return}~\KwSty{false}$\label{line:failhall}}
    \nl \lIf{$\left|A\right| = n$}{$(H,\,A,\,n) \gets (H \cup A,\,\emptyset,\,0)$\label{line:hall}}
}
\nl $\KwSty{return}~\KwSty{true}$ \;
}
\end{multicols}
\vspace*{0.4cm}% bug in algorithm2e + multicol: incorrect space left above the vertical line
\caption{An algorithm for the $k$-less subgraph isomorphism problem.}
\label{algorithm:sip}
\end{algorithm*}

\cref{algorithm:sip}. This is derived from a subgraph isomorphism due to
\citet{DBLP:conf/cp/McCreeshP15}, which in a recent evaluation of algorithm portfolios
\citep{thelionpaper} was the single strongest solver. As well as the modifications to solve $k$-less
and induced problems, we have made two simplifications: we do not use conflict-directed backjumping,
and we do not use iterated label filtering to recalculate neighbourhood degree sequences if vertices
are not present in any target domain after construction---in preliminary experiments, these
techniques were not beneficial for $k > 0$.

?? Discuss how we did wildcards, effectively unit domains. The all diff propagator.

?? Discuss how we do induced. We do not look at paths in the loop complements, since this is
expensive to calculate and provides little benefit in practice on the instances we considered.

@@

\begin{figure}[tb]
    \includegraphics*{gen-graph-runtimes.pdf}
    \caption{The cumulative number of instances solved over time, for the induced and non-induced
    problems, with different values of $k$. We also show the results of iteratively decreasing $k$
    until a solution is found, and in the induced case, the performance of a leading maximum common
    subgraph algorithm.}\label{figure:runtimes}
\end{figure}

@@

\begin{figure}[tb]
    \includegraphics*{gen-graph-which-k-by-family.pdf}
    \caption{The proportion of instances, in different families, which become satisfiable for
    increasing values of $k$. The larger instances are those where we can prove unsatisfiability for
    $k = 5$, whilst the gap between the top of the bar and the top of the graph is the fraction of
    instances where a timeout was reached for at least one value of $k$.}\label{figure:which-k}
\end{figure}

@@

We looked at . We saw that we could modify invariants, and that doing so still provided additional
domain reduction and filtering power.

Relaxed form of the problem which is easier to solve. Far from making maximum common subgraph scale
as well as subgraph isomorphism, but we can at least get somewhere.

Future work.

\section*{Acknowledgements}

We'd like to thank some people whose names we've removed for anonymous review for providing us with
source code and problem instances, and someone at one of our institutions for inspiration.
